---
title: "Swiftly squeaky clean: real-time adaptive management eradicates rat overpopulation and brings immediate benefits to tropical island biodiversity -- Data Analysis"
author: "Tati Micheletti"
date: "10/19/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document describes the data analysis done for the paper *Swiftly squeaky clean: real-time adaptive management eradicates rat overpopulation and brings immediate benefits to tropical island biodiversity*.  

## Load Libraries, source functions and set data directory

The first step is to load the libraries we need and source the functions that will prepare the data 
and run the models for each one of the species analyzed.

```{r libs}
if(!require("Require")){
    install.packages("Require")
}
library("Require")
Require("googledrive")
Require("reproducible")
Require("data.table")
Require("terra")
Require("unmarked")
Require("secr")
Require("tictoc")
Require("qs")
Require("ggplot2")
Require("zeehio/facetscales@archived")
# If any packages were installed here, restart your session and re-run the code!
# Here we list all functions and then source all of them:
allFls <- list.files(path = "./functions", full.names = TRUE)
for (fl in allFls){
  message(paste0("Sourcing the function ", basename(fl))) 
  source(file = fl) 
}
wd <- file.path(getwd(), "data")
```

# Wildlife species modeling

After loading libraries and sourcing needed functions, we load the data and 
format the information we need for the models. The 
`finalDataCOUNTS.csv` file contains all data for the endangered species. For all species, 
we used the N-mixture model described by Dail and Madsen (2011), which has been 
incorporated into the R package `unmarked`. This model, ran using the function 
`pcountOpen()`, allows for model formulations to identify population trends, 
with or without immigration, and for covariates on initial population (described 
as lambda in the model), population growth (gamma), observation probabilities 
(p) and immigration (iota), when this is considered. This model was chosen as it
could be used for all species, as it also allows for different distributions 
(i.e., Poisson, Negative Binomial and Zero Inflated). 

```{r loadAndCheckData}
Data <- loadData(dataLocation = file.path(wd, "finalDataCOUNTS.csv"))
weatherData <- loadWeatherData(dataLocation = file.path(wd, "weatherNoronhaMar17Nov23.csv"))
```


## Mabuia  

We started our analysis with the Mabuia (Noronha Skink; *Trachylepis atlantica*).   

For the mabuia, Negative Binomial Models would have been appropriate, as these model count data, 
especially if there is variability in the underlying rate of events. However, these are better 
suited for overdispersed data, which ours isn't. Mabuia data is, however, zero inflated. 
So ZIP models are more appropriate to use (see Kery, 
2018 and Stoklosa et al. 2022). Moreover, the mabuia population is closed to immigration, considering 
the island is not close enough to others to allow for the species to immigrate. As the predation rate 
by rats is high, we consider every sampling occasion as primary (i.e., open population). As we are 
mainly interested in observing the effect of eradication on population growth rate, we added a covariate 
for the time since the start of the eradication (TSE) for the model on Ilha do Meio, which could 
influence population growth (i.e., gamma). Mabuia observation was performed by three different 
researchers on both islands, and therefore, we also added researcherID as a covariate for detection. 
Lastly, we also added the possibility of landscape cover influencing the initial population size 
(i.e., lambda), as the number of individuals tends to be higher on densily vegetated areas (i.e., 
more space due to verticalization) than on open areas.  

```{r mabuiaMeio}
mabuiaMeio <- modelWildlife(DT = Data, 
                            island = "Ilha do Meio",
                            species = "Trachylepis atlantica",
                            weatherData = weatherData, 
                            useImmigration = FALSE, 
                            useRobustDesign = FALSE)

```

The best wildlife model for each species and island (i.e., Ilha Rata as Control, Ilha do Meio as Treatment) 
was then chosen based on Akaike Information Criterion (AIC). From the chosen model, 
a table was built showing species, island, cycle (matching Table XX), site, land cover type, parameter, 
estimate, standard error, lower confidence interval (LCI), upper 
confidence interval (UCI), model mixture, model formulation, AIC, K, and total model run time.

```{r mabuiaMeio2}
mabuiaMeioDT <- calculatePopulationChange(species = "Trachylepis atlantica",
                                          island = "Ilha do Meio",
                                          shortName = "Mabuia_Meio_",
                                          RobustDesign = FALSE)
```

We then did the same for the mabuia on Ilha Rata:    

```{r mabuiaRata}
mabuiaRata <- modelWildlife(DT = Data, island = "Ilha Rata",
                            species = "Trachylepis atlantica",
                            weatherData = weatherData,
                            useImmigration = FALSE, 
                            useRobustDesign = FALSE)

mabuiaRataDT <- calculatePopulationChange(species = "Trachylepis atlantica",
                                          island = "Ilha Rata",
                                          shortName = "Mabuia_Rata_",
                                          RobustDesign = FALSE)
```

## Elenia  

Next, we will continue our analysis with the Cucuruta (Noronha Elaenia; *Elaenia ridleyana*).  
The observation from Elaenias also rendered data with zero inflation. Therefore, 
we used a ZIP model. Our model also considers movement (iota) between the islands to be absent. 
Although there is evidence of a single individual flying from the archipelago to the Atol das Rocas,
approximately 145km away, the species has been observed in only 3 of the 26 islands and islets of 
Noronha. For this species, we used a robust design as we performed temporal replicates on the same 
day, considering a closed population within primary sampling occasions (sampling period within the 
same day), and open between primary sampling occasions (more than a day 
apart). Being an insectivore, we added the total amount of rainfall in the three months previous to 
the count as a covariate that could influence population growth, apart from TSE for Ilha do Meio. 
As the species was only observed in forest fragments and only by one researcher, neither landscape 
cover nor observerID were added as covariates. First we ran the models for Ilha do Meio:

```{r elaeniaMeio}
elaeniaMeio <- modelWildlife(DT = Data, island = "Ilha do Meio",
                             species = "Elaenia ridleyana",
                             weatherData = weatherData,
                             useImmigration = FALSE,
                             useRobustDesign = TRUE)

elaeniaMeioDT <- calculatePopulationChange(species = "Elaenia ridleyana",
                                          island = "Ilha do Meio",
                                          shortName = "Elaenia_Meio_",
                                          RobustDesign = TRUE)
```

And then for Ilha Rata:  

```{r elaeniaRata}
elaeniaRata <- modelWildlife(DT = Data, island = "Ilha Rata",
                             species = "Elaenia ridleyana",
                             weatherData = weatherData,
                             useImmigration = FALSE,
                             useRobustDesign = TRUE)

elaeniaRataDT <- calculatePopulationChange(species = "Elaenia ridleyana",
                                          island = "Ilha Rata",
                                          shortName = "Elaenia_Rata_",
                                          RobustDesign = TRUE)
```

## Land Crab  

The analysis of land crabs (*Johngarthia lagostoma*) was then performed. This 
species is known to present increase movement during breeding season. Lesser 
wave action associated with neap high tides (in first and last quarters of moon 
cycle) seem to be preferred. In the same way, peak periods of rain (i.e., March 
or April) also seem to be important factors for this species' movement. Therefore, we 
added the moon phase and total rainfall in the three previous months as covariates influencing 
population growth and observation, apart from apart from TSE for Ilha do Meio. Crab data is also 
zero inflated, so we tested both ZIP and NB models. Similarly to Mabuia, although 
the NB model presented a smaller AIC, the dispersion parameter was not significant, 
while the zero inflation was in the ZIP model. Therefore, we opted for a ZIP model. It is important 
to notice that a four models for this species on Ilha do Meio did not converge throwing the error 
`Hessian is singular.`. All of these models had landscape cover as a covariate for lambda and 
TSE as covariate for gamma. This warning generally happens when parameters are non-identifiable and  
can be usually resolved by either simplifying the model, or providing good starting values. 
Even after trying to provide starting values for the models, however, these still did not converge. 
This suggests the model formulation cannot capture the signal correctly and the estimates are likely 
unreliable. These are, however, still reported in Table X with a footnote.

```{r crabMeio}
crabMeio <- modelWildlife(DT = Data, island = "Ilha do Meio", 
                              species = "Johngarthia lagostoma",
                             weatherData = weatherData,
                             useImmigration = FALSE,
                             useRobustDesign = FALSE)

crabMeioDT <- calculatePopulationChange(species = "Johngarthia lagostoma",
                                          island = "Ilha do Meio",
                                          shortName = "Crab_Meio_",
                                          RobustDesign = TRUE)
```

After modelling crabs for the Ilha do Meio, we perform the same analysis for Ilha Rata:

```{r crabRata}
crabRata <- modelWildlife(DT = Data, island = "Ilha Rata", 
                              species = "Johngarthia lagostoma",
                             weatherData = weatherData,
                             useImmigration = FALSE,
                             useRobustDesign = FALSE)

crabRataDT <- calculatePopulationChange(species = "Johngarthia lagostoma",
                                          island = "Ilha Rata",
                                          shortName = "Crab_Rata_",
                                          RobustDesign = TRUE)
```

## Masked Booby

To perform the analysis for a seabird species, the Atoba Mascarado (masked booby; 
*Sula dactylatra*), surveys were performed in all breeding sites across the Ilha 
do Meio (n = 7), and on the only breeding site on Ilha Rata. We used adult counts 
as a proxy for breeding site quality. As this species does not permanently stay on the island, we 
included immigation and modelled it in function of TSE, in the same way for gamma for the Ilha do 
Meio. The only model ran for Ilha Rata was the constant one.

```{r crabMeio}
maskedBoobyMeio <- modelWildlife(DT = Data, island = "Ilha do Meio", 
                                     species = "Sula dactylatra",
                             weatherData = weatherData,
                             useImmigration = TRUE,
                             useRobustDesign = FALSE)
maskedBoobyMeioDT <- calculatePopulationChange(species = "Sula dactylatra",
                                          island = "Ilha do Meio",
                                          shortName = "maskedBooby_Meio_",
                                          RobustDesign = TRUE)
```

After modelling crabs for the Ilha do Meio, we perform the same analysis for Ilha Rata:

```{r crabRata}
maskedBoobyRata <- modelWildlife(DT = Data, island = "Ilha Rata", 
                                     species = "Sula dactylatra",
                             weatherData = weatherData,
                             useImmigration = TRUE,
                             useRobustDesign = FALSE)
maskedBoobyRataDT <- calculatePopulationChange(species = "Sula dactylatra",
                                          island = "Ilha Rata",
                                          shortName = "maskedBooby_Rata_",
                                          RobustDesign = TRUE)

```

We then joined all species tables into one and calculate, for each species, the differential 
population growth index to be plotted.

```{r popChangeTable}
finalTableName <- file.path("outputs/finalTable.csv")
if (!file.exists(finalTableName)){
 finalTable <- rbindlist(list(
   mabuiaMeioDT,
   mabuiaRataDT,
   elaeniaMeioDT,
   elaeniaRataDT,
   crabMeioDT,
   crabRataDT,
   maskedBoobyMeioDT,
   maskedBoobyRataDT), use.names = TRUE)
  write.csv(tb, finalTableName)
} else {
  finalTable <- data.table::fread(finalTableName)
}
```

For completeness, we also provide all model formulation's tested for all species and their parameter 
estimates on Table X. 

```{r allModelsTable}

fullModTable <- createFullModelTable(finalTable = finalTable)

```



# Rat population modeling

We also modeled rat population during the eradication efforts. For that, we used spatially 
explicit capture-recapture techniques and the R package `secr` for analysis. Each 
trapping sampling session data (i.e., campaign data of consecutive days) was 
used as a stand-alone, independent dataset. This was done due to the adjustments 
needed on number and disposition of traps for each new monitoring campaign, as 
described in the manuscript. As rats were only captured in the first two campaigns, 
despite increase trapping effort on the third campaign, only these two were modeled.  
The function `getRatDensity()` performs all necessary steps, with comments.

``` {r getRatD}
blackRatData <- getRatDensity()
```

We tested different model formulations (i.e., landscape covariates and also 
different buffers; see previous commits to see code). Model diagnostics (i.e., AIC) supported 
the use of the simplest model (i.e., detection probability doesn't change), and the island 
shapefile as a mask. Applying the mask or buffer of 200 didn't change the results, but other buffers 
did.

# Building final plot

At last, we calculated the Differential Population Growth Index: This index represents the 
difference in predicted population growth rates between Ilha do Meio and Ilha Rata for each 
species. By subtracting the predictions for Ilha Rata from those of Ilha do Meio, the index 
aims to isolate the specific factors influencing population growth on Ilha do Meio, independent of 
regional drivers that could affect population growth uniformly across both islands. This adjustment 
helps to eliminate the influence of broader regional factors like extreme events and natural 
population variation cycles, allowing for a more focused analysis of the drivers specific to 
Ilha do Meio. We then plotted the results of all species (i.e., black rat density, and Differential 
Population Growth Index of mabuias, crabs, elaenia and masked booby) in the same plot, Figure X.  

At last, we build a plot to show the difference in growth rate through time in function 
of time since eradication (TSE) using the control to control for seasonality. We use growth 
rate instead of other parameters (i.e., population density) due to the compounded uncertainty 
in the last due to both uncertainty in initial population and population growth rate.  
Moreover, growth rate shows exactly what we sought with the current work, the population trend 
through time.

```{r plotPopGrowth}
plotGrowth <- plotPopulationGrowth(wildlifeDataset = finalTable, 
                                   whatToPlot = "populationGrowthRate",
                                   ratsDataset = blackRatData)
```


